---
title: "GO analysis"
author: "Jason T. Serviss"
date: "09/06/2017"
output:
  html_document:
    code_folding: hide
    highlight: pygments
    theme: readable
---

<style> 
.col2 { 
    columns: 2 200px; /* number of columns and width in pixels*/ 
    -webkit-columns: 2 200px; /* chrome, safari */ 
    -moz-columns: 2 200px; /* firefox */ 
} 
.caption { 
    color: FireBrick; 
    font-weight: bold; 
    font-size: 0.75em; 
}
</style>


```{r libraries, message=FALSE}
packages <- c(
  "acidAdaptedRNAseq",
  "GO.db",
  "GOSim",
  "igraph",
  "ggplot2",
  "ggraph",
  "multipanelfigure",
  "grid",
  "gtable",
  "magrittr",
  "printr",
  "tibble",
  "dplyr",
  "stringr",
  "purrr",
  "tidyr"
)
purrr::walk(packages, library, character.only = TRUE)
rm(packages)

#source('~/Github/diverseScripts/R/scripts/largeDivergingPalettes.R')
```

Previously we have performed a gene ontology (GO) term enrichment analysis using 
the differentially expressed genes (adjusted p-value < 0.05). Due to the fact 
that many of the significant GO terms are related to similar biological 
processes, it is desireable to facilitate the understanding of the results to 
combine similar terms into groups, thus minimizing the number of terms to be 
simultaniously considered. 

In this analysis we use comminuty detection in order to reduce terms into groups
but still strive to maintain the majority of the information concerning the 
relationship of GO terms to each other. In addition, we highlight the 
differences in gene expression between the experimental conditions in the 
detected communities. Finally, we show the specific expression of a subset of 
the genes in some of the GO terms in several of the communities.

```{r, message=FALSE}
#load GO result and subset signignificant terms
data(topGOresult, package = "acidAdaptedRNAseq")
sig <- topGOresult[topGOresult$classicFisher < 0.05, ]
```

To understand the relationship between the terms found to be significant in the 
GO term enrichment analysis, we first download the GO graph for all of the terms.
This returns a graph of all the significant terms as well as their ancestors. 

We then proceed to perform community detection and collapse the graph and 
simplify the edges. For the community detection we use the spin-glass algorithm.
From the available algorithms in the igraph package, a subset of these are not 
appropriate for directed graphs and these were not evaluated. Those which are 
appropriate for directed graphs were all tested and evaluated individually by a)
the number of communities detected and b) the biological relevance of the 
communities detected. a) was desired to be the smallest number possible without 
have a strong negative impact on b). b) was evaluated manually by reviewing the
GO terms present in each commuity and their biological relationship to each
other. Although both the infomap and spinglass community detection method 
performed well for a), it was determined that the spinglass algorithm performed
better for b) and was therefore implemented in the analysis.

A resonably detailed summary of the community detection algorithms available in 
the igraph package can be found at:
https://stackoverflow.com/questions/9471906/what-are-the-differences-between-community-detection-algorithms-in-igraph

Collapsing the graph on communities results in each community being represented 
by one node. Furthermore, in each detected community the node with the highest 
authority score is determined. Due to the orientation of the graph, this node 
represents the term in the community that has the most incomming edges, also 
known as the node with the highest indegree, and is thus the node closest to the
root of the graph (i.e. the "all" term). In the downstream analysis, the node 
with the highest authority score is used to represent the community.
In addition, in the case where there were edges between the terms now 
located in communities, a single edge between the communities that those terms 
are now located in, is used to represent this relationship.

```{r, message=FALSE}
#get GO graph
g <- downloadGOgraph(sig$GO.ID)

#save(g, file = "./goGraph.rda", compress = "bzip2")

#run community analysis
tmp <- collapseGraph(g)
cg <- tmp[[1]]
comSummary <- tmp[[2]] %>%
    left_join(dplyr::select(sig, GO.ID, classicFisher), by=c("GOID" = "GO.ID"))
```

The community detection algorithm detected a total of 39 communities. Below we 
calculate and show the number of genes included in each community as an 
indication of the communities information content. The results indicate that the
smallest community is comprised of only 5 genes (lymphocyte apoptotic process) 
whereas the largest community (cellular metabolic process) has just under 30000.

```{r, message = FALSE}
#calculate genes per term
uCommunityNames <- unique(comSummary$communityName)
genesPerCom <- lapply(1:length(uCommunityNames), function(x) {
    bool <- comSummary$communityName == uCommunityNames[x]
    idsToGet <- comSummary[bool, ]$GOID
    genesForIdsString <- sig[sig$GO.ID %in% idsToGet, "ID"]
    genesForIdsList <- strsplit(genesForIdsString, ", ")
    length(unlist(genesForIdsList))
})
names(genesPerCom) <- uCommunityNames

namedListToTibble(genesPerCom) %>%
    setNames(c("communityName", "nrCommunityGenes")) %>%
    inner_join(distinct(select(comSummary, communityName, communityTerm))) %>%
    select(communityTerm, nrCommunityGenes, -communityName)
```

As a measure of the effectivity of the community detection we can review the 
dissimilarity between communities by first calculating the Jaccard similarity
between communities and then performing hierarchical clustering using 
1-similarity. The results (below) indicate that the majority of communities have
a reasonable level of dissimilarity with the exception of the 
*protein refolding* and *antigen processing and presentation of exogenous...*
community.
```{r, fig.align='center', fig.height=17, fig.width=15, eval=TRUE, message=FALSE, warning=FALSE}
cgSim <- similarity(cg, mode = "all", method = "jaccard")
colnames(cgSim) <- unique(
    comSummary[match(V(cg)$name, comSummary$communityID), ]$communityTerm
)
rownames(cgSim) <- unique(
    comSummary[match(V(cg)$name, comSummary$communityID), ]$communityTerm
)
heatmap(1 - cgSim, margins = c(20,20))
```

```{r}
#adds attributes to the collapsed graph
cg <- addAttributes1(cg, comSummary)
```

To give a graphical representation of the size of each community, we now add 
all of the terms found to be significant in the GO term enrichment analysis 
back to the collapsed graph. In addition, we draw an edge from each individual 
term to the community that is is a member of. 

It should be noted that now the graph only contains terms found to be 
significant in the GO term enrichment analysis with the exception of the 
community nodes. Due to the fact that the authority score on the whole graph 
was used to determine the community nodes, it can be the case that the community 
node itself was not actually found to be a significant term. Despite this, each
of the terms in the community were significantly enriched. Therefore the
community nodes should be viewed as a summary or label encasing the significant
terms rather than a significant term itself (although in some cases it is).

```{r, fig.align='center', fig.height=17, fig.width=15, eval=TRUE, message=FALSE, warning=FALSE}
#add vertices and edges from uncollapsed graph
toAdd <- comSummary %>%
    filter(!GOID %in% communityID) %>%
    filter(classicFisher < 0.05) %>%
    select(GOID, communityID) %>%
    setNames(c("from", "to"))

cg <- addBackVertexAndEdges(cg, toAdd)

#add a line break in long community groups so the plot legend can be seen
comSummary$communityTerm <- ifelse(
    comSummary$communityTerm == "antigen processing and presentation of exogenous peptide antigen via MHC class I, TAP-independent", 
    "antigen processing and presentation of exogenous\npeptide antigen via MHC class I, TAP-independent",
    comSummary$communityTerm
)

comSummary$communityTerm <- ifelse(
    comSummary$communityTerm == "somatic recombination of immunoglobulin genes involved in immune response", 
    "somatic recombination of immunoglobulin\ngenes involved in immune response",
    comSummary$communityTerm
)

comSummary$communityTerm <- ifelse(
    comSummary$communityTerm == "generation of precursor metabolites and energy", 
    "generation of precursor\nmetabolites and energy",
    comSummary$communityTerm
)

#add attributes
cg <- addAttributes2(cg, comSummary)

#plot
plotCollapsedGraph(cg, 12, 3)
collapsed <- plotCollapsedGraph(cg)
mylegend <- g_legend(collapsed)
collapsed <- collapsed + theme(legend.position = "none")
```
<div class=caption>
A network graph visualizing the results from the community analysis. Edge (line)
direction is represented by color with edges originating from a node inheriting 
that nodes color. Edges between community nodes (large points) indicate that the
GO terms representing the nodes are each other’s ancestors or offspring 
dependent on the direction of the edge. Edges between term nodes (small points) 
and community nodes indicate the terms inclusion in that community. 
</div>

In summary, many of the detected communities represent processes which we would 
expect to be differentially regulated in the experimental conditions (i.e. cell
cycle, cell death, cell adhesion). In addition, multiple communities related
to metabolism can be identified (i.e. cellular metabolic process, carboxylic
acid metabolic process, lipid metabolic process).

It should be noted that some edges that could be expected may not appear. For 
example, one may expect an edge between cell cycle and cell proliferation, due
to the fact that they are biologically related. In this case, and potentially 
others, the terms are actually connected via an intermediate term 
(single-orgamism cellular process) and therefore do not have an edge directly 
connecting the nodes. This is unfortunate for the understanding of the results 
but has, so far, proved to be unavoidable due to the nature of the analysis.

Due to the fact that the GO term enrichment analysis was performed using 
significant (alpha < 0.05) differences in gene expression between experimental
conditions, we would expect the gene expression in the terms found to be 
significant (and therefore represented in the graph above) to be different. 
Despite this, a visualization of this may help to strengthen the belief that the
communities in the graph represent true differences in the experimental 
conditions. To facilitate this understanding we plot the differences in 
each experimental condition per community below.

```{r, fig.align='center', fig.height=17, fig.width=15, eval=TRUE, message=FALSE, warning=FALSE}
#extract all gene IDs per GO term
genePerTerm <- extractGenesFromGO()

#calculate z-score 
zNorm <- rescale()

#merge data, note that GO graph terms not in sig are included in comSummary
heatmap <- comSummary %>%
    filter(!is.na(classicFisher)) %>%
    full_join(genePerTerm, by = "GOID") %>%
    left_join(zNorm, by = "ID") %>%
    gather(condition, zScore, -communityName, 
           -communityID, -communitySize, -communityTerm,
           -GOID, -authority.score, -Term, 
           -classicFisher, -ID
    )

#run clustering per community to get row order
order <- clusterForOrder(heatmap)

#merge order into summary
heatmap <- left_join(heatmap, order, by = c("communityName", "ID"))

#plot
heatmap$plotCondition <- plotCondRename(heatmap$condition)
heatmap <- plotHeatmap(heatmap)
```

<div class=caption>
A heatmap representing gene expression profiles in the detected communities. 
Communities are indicated by the color bar on the right side of the heatmap. 
Z-score indicates the regularized log transformed expression values scaled 
between 0 and 1.
</div>

Finally, we wish to highlight genes which would be expected to be disregulated
in the experimental model and their location in the community graph. To do this,
we extract the expression values for individual genes and plot them in each 
experimental condition while connecting them each back to the term and community
they are located in.

```{r}
#setup rld expression values for merge
rld <- deResultsRld %>%
    as.data.frame() %>%
    rownames_to_column() %>%
    dplyr::rename(ID = rowname) %>%
    as_tibble() %>%
    gather(condition, expression, -ID)

#set up gene p-values for merge
ps <- deResults %>%
    dplyr::select(ID, padj) %>%
    as_tibble()

#setup data for plot
barDat <- comSummary %>%
    filter(!is.na(classicFisher)) %>%
    full_join(genePerTerm, by = "GOID") %>%
    left_join(rld, by = "ID") %>%
    left_join(geneAnnotation, by = "ID") %>%
    left_join(ps, by="ID")

barDat$plotCondition <- plotCondRename(barDat$condition, reps = FALSE)

#Specify terms to plot
terms <- c(
    "positive regulation of autophagy",
    "response to osmotic stress",
    "cell death",
    "positive regulation of metabolic process",
    "negative regulation of cell differentiation",
    "regulation of protein ubiquitination",
    "cell cycle",
    "cell proliferation"
)
```

<div class=col2>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
#plot for rmarkdown
plotGenes(barDat, terms[1])
```

<br></br>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
plotGenes(barDat, terms[2])
```

<br></br>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
plotGenes(barDat, terms[3])
```

<br></br>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
plotGenes(barDat, terms[4])
```

<br></br>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
plotGenes(barDat, terms[5])
```

<br></br>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
plotGenes(barDat, terms[6])
```

<br></br>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
plotGenes(barDat, terms[7])
```

<br></br>

```{r, fig.align='center', fig.height=8, fig.width=7, eval=TRUE, message=FALSE, warning=FALSE}
plotGenes(barDat, terms[8])
```

<br></br>
</div>
<div class=caption>
Gene expression profiles for selected terms in a subset of the detected 
communities for parental and acid adapted cells.
</div>

```{r, message=FALSE}
#plot for figure
p1 <- plotGenes(barDat, terms[1]) 
mylegend2 <- g_legend(p1)

p1 <- plotGenes(barDat, terms[1]) + theme(legend.position = "none")
p2 <- plotGenes(barDat, terms[2]) + theme(legend.position = "none")
p3 <- plotGenes(barDat, terms[3]) + theme(legend.position = "none")
p4 <- plotGenes(barDat, terms[4]) + theme(legend.position = "none")
p5 <- plotGenes(barDat, terms[5]) + theme(legend.position = "none")
p6 <- plotGenes(barDat, terms[6]) + theme(legend.position = "none")
p7 <- plotGenes(barDat, terms[7]) + theme(legend.position = "none")
p8 <- plotGenes(barDat, terms[8]) + theme(legend.position = "none")

#set up figure panels
figure1 <- multi_panel_figure(
  width = c(200, 100, 100, 100),
  height = 350,
  rows = 5,
  unit = "mm"
 )

#plot to panels
figure1 %<>% fill_panel(mylegend, column = 1:4, row = 5, label = "")
figure1 %<>% fill_panel(collapsed, column = 1, row = 1:4, label = "A")
figure1 %<>% fill_panel(heatmap, column = 2, row = 1:4, label = "B")
figure1 %<>% fill_panel(p1, column = 3, row = 1, label = "C")
figure1 %<>% fill_panel(p2, column = 3, row = 2, label = "D")
figure1 %<>% fill_panel(p3, column = 3, row = 3, label = "E")
figure1 %<>% fill_panel(p4, column = 3, row = 4, label = "F")
figure1 %<>% fill_panel(p5, column = 4, row = 1, label = "G")
figure1 %<>% fill_panel(p6, column = 4, row = 2, label = "H")
figure1 %<>% fill_panel(p7, column = 4, row = 3, label = "I")
figure1 %<>% fill_panel(p8, column = 4, row = 4, label = "J")

#figure1

#save final figure
ggsave(
    plot = figure1,
    './figure.pdf',
    width = 550,
    height = 350,
    units = "mm",
    limitsize = FALSE
)
ggsave(
    plot = mylegend2,
    './legend.pdf'
)
print(getwd())
```

```{r}
sessionInfo()
```
