---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

To repete the total analysis you must run the following scripts pre-installation from the package root directory.
Rscript -e "source('./inst/R.scripts/getAnnotation.R'); getAnnotation()"
Rscript -e "source('./inst/data-raw/processGO.R'); processGO()"
Rscript -e "source('./inst/data-raw/runMe.R'); runMe()"
Rscript -e "source('./inst/data-raw/runHCT.R'); runHCT()"
##add correlation script and note RAM requirements
after installation:

To re-create the MA plot from DESeq2 use: 
```{r}
data(HCT116.DESeq2.Obj) ##or any DESeq2 object
plotMA(results(HCT116.DESeq2.Obj))
```

To re-create the mean variance trend plot from limma/voom use:

```{r}
data(Me.limma.Obj)
targets <- data.frame(
        library <- c("parentalA","parentalB","parentalC","adaptedA","adaptedB","adaptedC"),
        condition <- factor(c(rep("parental", 3),rep("adapted", 3)),levels=c("parental", "adapted"))
)

design <- DeMilitoRNAseq:::.design.contrasts.limma(targets)
hsa.voom <- limma::voom(Me.limma.Obj, design=design, plot=TRUE)
```